---
title: "Explore And Summarize_Wine"
author: "Maitrayee Ranade Bordikar"
date: "July 13, 2017"
output: html_document
html_document:
    fig_width: 6
    fig_height: 4
bibliography: library.bib
---

========================================================

This report is on exploration of the wine qualities data set both Whites and 
Reds [@P.CortezA.CerdeiraF.Almeida2009]. Below is information about the datasets
from the Authors.

The two datasets are related to red and white variants of the Portuguese 
"Vinho Verde" wine. For more details, consult: http://www.vinhoverde.pt/en/ or 
the reference [@P.CortezA.CerdeiraF.Almeida2009]. Due to privacy and logistic 
issues, only physicochemical (inputs) and sensory (the output) variables are 
available (e.g. there is no data about grape types, wine brand, wine selling 
price, etc.).

These datasets can be viewed as classification or regression tasks. The classes 
are ordered and not balanced (e.g. there are munch more normal wines than
excellent or poor ones). Outlier detection algorithms could be used to detect 
the few excellent or poor wines. Also, we are not sure if all input variables 
are relevant. So it could be interesting to test feature selection methods. 

Number of Instances: red wine - 1599; white wine - 4898. 

Number of Attributes: 11 + output attribute
  
Description of attributes:

   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
   (do not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of
   levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' 
   and flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
   growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
   concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the 
   percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very 
   acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
   (S02) levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine

   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)


Since both the datasets have same set of variables, I am combinging the two sets
and creating a new attribute of color to differentiate. The analysis could show
some ineteresting results while comaparing Red and White wines.

The code below loads all the libraries needed for the analysis.
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(scales)
library(corrplot)

library(randomForest)
library(party)

#library(reprtree)


```

Loading dataset below, combing it to form one dataframe. Creating another 
variable of color to identify red or white wine.

```{r echo=FALSE, Load_the_Data}
# Load the Data
wqw = read.csv("D:\\Maitrayee\\1DataScience\\Udacity\\P4\\wineQualityWhites.csv")
wqr = read.csv("D:\\Maitrayee\\1DataScience\\Udacity\\P4\\wineQualityReds.csv")

wqw$color <- 'White'
wqr$color <- 'Red'


wq <- rbind(wqw, wqr)

```

Data Summary:
```{r echo=FALSE, Summary_Data}
str(wq)
```
The combined datsett now has 12 input variables and 1 output variables. And a 
total of 6497 observations.

This report is divided into explroring single variables at a time, then two 
variables followed by multiple variables. 

# Univariate Plots Section

Since I will be looking at the variables of Red and White wines simultaneosly, I 
am using the multiplot function[@Chang, Winston] to print plots in grid format.

```{r echo= FALSE, Fucntions }
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}


cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat) 
  p.mat
}

```

Looing at the quality of wines in the database.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_1}

ggplot(data=wq,  aes(x=quality, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 1, alpha = .5, color = 'black') + 
  # ggtitle("Quality") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Quality") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

m <- mean(wq$quality[wq$color == 'White'])
m2 <- mean(wq$quality[wq$color == 'Red'])

```

The wine quality ratings is normally distributed with its mean equal to 5.87 and
5.63 for White and Red wines respetively. Next, taking alook at the input 
variables that could determine the quality ratings of wine begining with alcohol
and pH.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_2}

p1 <- ggplot(data=wq,  aes(x=pH, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.1, alpha = .5, color = 'black') + 
  #ggtitle("pH") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("pH") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

p2 <- ggplot(data=wq,  aes(x=alcohol, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.5, alpha = .5, color = 'black') + 
  #ggtitle("Alcohol") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Alcohol") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

multiplot(p1, p2, cols = 1)

```

Both pH and alcohol levels are normally disctributed but alcohol distribution is
slightly skewed. pH levels for both red and white wine is between 2.75-3.75. 
Alcohol percentage for both most of the White and Red wines is between 9% and 11
%. 

Next looking at the different acidities.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_3}

p1 <- ggplot(data=wq, aes(x=fixed.acidity, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.5, alpha = .5, color = 'black') + 
  #ggtitle("Fixed Acidity") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Fixed Acidity") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))
  
p2 <- ggplot(data=wq, aes(x=citric.acid, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.05, alpha = .5, color = 'black') + 
  #ggtitle("Citric Acid") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Citric Acid") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

p3 <- ggplot(data=wq, aes(x=volatile.acidity, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.05, alpha = .5, color = 'black') + 
  #ggtitle("Volatile Acidity") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Volatile Acidity") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

multiplot(p1, p2, p3, layout = matrix(c(1,2,3), nrow=3, 
                                               byrow=TRUE))
  
  
```


For white wines both the acidity levels and citric acid is normally distributed 
with mean and median comparable to each other. For red wines the citric acid
level distribution is more skewed with the peak at 0. Could this be because
Red wines are considered sweeter compared to White wines?

Next looking at the residual sugar.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_4}

ggplot(data=wq,  aes(x=residual.sugar, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 1, alpha = .5, color = 'black') + 
  #ggtitle("Residual Sugar") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Residual Sugar") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))
  
```

The distribution of residual sugar is skewed with most of Red and White wines 
having sugar less than 5. Red wines appear to have more samples with sugar less
than 5. It will be interesting to see how the quality of the wine is affected by 
the sweetness of the wine. 

Residual Sugar and Citric acid has skewed distribution, so looking at them in 
box plots.

``` {r echo=FALSE, warning=FALSE, Univariate_plot_5}

p1 <- ggplot(wqr, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

p2 <- ggplot(wqw, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'blue' ) +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

multiplot(p1,p2, cols = 2)


p1 <- ggplot(wqr, aes( x = 1, y = citric.acid ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

p2 <- ggplot(wqw, aes( x = 1, y = citric.acid ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'blue' ) +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

multiplot(p1,p2, cols = 2)
```


Red wine has many outliers for residual sugar, while white wine has many 
outliers for citric acid. Looking at density of both the wine types, with 
histogram and boxplot.


``` {r echo=FALSE, warning=FALSE, Univariate_plot_6}


ggplot(data=wq,  aes(x=density, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.001, alpha = .5, 
                 color = 'black') + 
  # ggtitle("") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Density") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

p1 <- ggplot(wqr, aes( x = 1, y = density ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

p2 <- ggplot(wqw, aes( x = 1, y = density ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'blue' )  +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
          stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

multiplot(p1,p2, cols = 2)



```


The mean of density for both wine types is comaprable but Red wine has more 
outliers. Next looking at sulphates and chlorides of the wines.

```{r echo=FALSE, warning=FALSE, Univariate_Plots_7}

p1 <- ggplot(data=wq,  aes(x=chlorides, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.005, alpha = .5, color = 'black') + 
 # ggtitle("Chlorides") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Chlorides") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

p2 <- ggplot(data=wq,  aes(x=sulphates, fill = color)) + 
  geom_histogram(position = "dodge", binwidth= 0.05, alpha = .5, color = 'black') + 
 # ggtitle("sulphates") + theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Sulphates") + ylab("Count") +
    scale_fill_manual(values=c("red", "white"))

multiplot(p1, p2, cols = 1)

m1 <- mean(wq$chlorides[wq$color == 'Red'])
m2 <- mean(wq$chlorides[wq$color == 'White'])

m3 <- mean(wq$sulphates[wq$color == 'Red'])
m4 <- mean(wq$sulphates[wq$color == 'White'])
  
```


For both Red and White wines the cholrides and sulphates distribution is normal
but its mean is shifted with respect to eah other. Red's chloride mean 
(0.087) is higher than White's mean (0.046). And Red's sulphates mean (0.66) is 
also higher than White's (0.49).


# Univariate Analysis



### What is the structure of your dataset?
There are 13 variables and 4898 observations in this database. All the variables
have numeric values, so all the visuallizations have been in histogram format. 
The quality of the wine is a number between 0-10. In this samples set quality is
normally distributed with mean of 5.85 and median of 6.

### What is/are the main feature(s) of interest in your dataset?
The main feature I would like to ivestigate is how the quality of wines is 
determined by its other features such as acidity, sugar, chlorides and sulphates
etc.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Note that all the features have numeric values and not 
factors. Their inter-relation and its effect on the wine quality will be 
intersting to explore.

### Did you create any new variables from existing variables in the dataset?
This dataset is a tidy data set so I did not have to perform any data wrangling.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

All the variables in the data set have a numeric value and not factors. It will 
be interesting to explore the variables (quality, alcohol) by factorizing them 
and checking its dependancy on other variables. This analysis is in the next 
section.




# Bivariate Plots Section


Factorizing wine quality into 3 categories poor (quality <= 5.2), fine (5.2 > 
quality <=7.2) and excellent (quality > 7.2). Also, factorizing alcohol (%) into
5 categories light (alcohol(%) <= 10.0), medium-low (10.0 < alcohol(%) <= 11.5),
medium (11.5 < alcohol(%) <= 13.5), medium-high (13.5 < alcohol(%) <= 15) and 
high (alcohol(%) >15).


```{r echo=FALSE, warning=FALSE, Bivariate_Plots_1}

  wq$qualityF <- 'poor'
  wq$qualityF[wq$quality >= 5.2 & wq$quality <=7.2] <- 'fine'
  wq$qualityF[wq$quality > 7.2] <- 'excellent'
  wq$qualityF <- factor(wq$qualityF)
  
  wq$qualityF_sn <- 'p'
  wq$qualityF_sn[wq$quality >= 5.2 & wq$quality <=7.2] <- 'f'
  wq$qualityF_sn[wq$quality > 7.2] <- 'e'
  wq$qualityF_sn <- factor(wq$qualityF_sn)
  
  wq$alcoholF <- 'low'
  #wq$alcoholF[wq$alcohol <= 10.0] <- 'low'
  wq$alcoholF[wq$alcohol > 10.0 & wq$alcohol <=11.5] <- 'medium-low'
  wq$alcoholF[wq$alcohol > 11.5 & wq$alcohol <=13.5] <- 'medium'
  wq$alcoholF[wq$alcohol > 13.5 & wq$alcohol <=15] <- 'medium-high'
  wq$alcoholF[wq$alcohol > 15] <- 'high'
  wq$alcoholF <- factor(wq$alcoholF)
  
  
  p1 <- ggplot(data=wq,  aes(x=qualityF, fill = color)) + 
    geom_bar(position = "dodge", alpha = .5, color = 'black') + 
    #ggtitle("Quality") + theme(plot.title = element_text(hjust = 0.5)) +
    xlab("Quality (Factors)") + ylab("Count") +
    scale_fill_manual(values=c("red", "white")) +
    scale_x_discrete(limits=c('poor','fine', 'excellent'))

 p2 <- ggplot(data=wq,  aes(x=alcoholF, fill = color)) + 
    geom_bar(position = "dodge", alpha = .5, color = 'black') + 
    #ggtitle("Alcohol") + theme(plot.title = element_text(hjust = 0.5)) +
    xlab("Alcohl (Factors)") + ylab("Count") +
    scale_fill_manual(values=c("red", "white")) +
    scale_x_discrete(limits=c('low','medium-low', 'medium', 'medium-high', 'high'))
 
multiplot(p1, p2, cols = 1)
  

```

Plotting the correlation coefficient between variables for each type of wine. 
Created a separate subset for red and white wine with just the input and output
variables. Using the corrplot library to plot the correlation matrix.
[@Kassambra]

``` {r, fig.width=10, fig.height=11, echo = FALSE, warning= FALSE, Bivariate_Plots_2}

wqr_corr <- subset(wq, color == "Red",select= c(quality, fixed.acidity, 
               volatile.acidity, citric.acid, residual.sugar, chlorides, 
               free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates,
               alcohol))

wqw_corr <- subset(wq, color == "White",select= c(quality, fixed.acidity, 
               volatile.acidity, citric.acid, residual.sugar, chlorides, 
               free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates,
               alcohol))

M_r <- cor(wqr_corr)
p.mat_r <- cor.mtest(M_r)

M_w <- cor(wqw_corr)
p.mat_w <- cor.mtest(M_w)

title <- "Red Wine - correlation"
corrplot(M_r, method="color", 
         diag=FALSE, # tl.pos="d",
         type="upper",  
     title=title, 
         addCoef.col = "black", # Add coefficient of correlation
         # Combine with significance
         p.mat = p.mat_r, sig.level = 0.05, insig = "blank", 
         # hide correlation coefficient on the principal diagonal
     mar=c(0,0,1,0) # http://stackoverflow.com/a/14754408/54964
         )

title <- "White Wine - correlation"
corrplot(M_w, method="color", 
         diag=FALSE, # tl.pos="d",
         type="upper",  order="hclust", 
     title=title, 
         addCoef.col = "black", # Add coefficient of correlation
         # Combine with significance
         p.mat = p.mat_w, sig.level = 0.05, insig = "blank" ,
         # hide correlation coefficient on the principal diagonal
     mar=c(0,0,1,0) # http://stackoverflow.com/a/14754408/54964
         )


```

The correlation matrix shows only the significant correlation coefficient 
between all the variables for each of the wine types separately. For both Red 
and White wine the quality of wine has positive correlation with alochol of 
approx 0.45. Red's quality also has a significant correlation of approx - 0.4 
with its volatile acidity. While whites' quality has significant correlation 
with  density, cholrides and total-sulphur-dioxide of -0.31, -0.21 and -0.17 
respetively.

Exploring alcohol and quality relation.

``` {r echo=FALSE, warning=FALSE, Bivariate_plot_3}


p1 <- ggplot(aes(qualityF, 
            alcohol), 
        data = subset(wq, color == "Red",select= c(qualityF, alcohol)) )+
  geom_jitter( alpha = .3, color = 'blue')  +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Quality - Factored") + ylab("Alcohol [%]") + 
  scale_x_discrete(limits=c('poor','fine', 'excellent')) +
  geom_boxplot( alpha = .5,color = 'red')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

p2 <- ggplot(aes(qualityF, 
            alcohol), 
        data = subset(wq, color == "White",select= c(qualityF, alcohol)) ) +
  geom_jitter( alpha = .3, color = 'blue')  +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Quality- Factored") + ylab("Alcohol [%]") + 
  scale_x_discrete(limits=c('poor','fine', 'excellent')) +
  geom_boxplot( alpha = .5,color = 'black')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

multiplot(p1,p2, cols = 2)


```

Both wine types have more fine wines. Alcohol percentage is less in poor wines 
and it increases with wine quality. Although, poor wine has more outliers with 
higher alcohol percentage levels. The mean of alcohol for each wine type and 
quality is comparable. Next looking at the relation between density and alcohol 
for each wines.

``` {r echo=FALSE, warning=FALSE, Bivariate_plot_4}

 ggplot(aes(x = alcohol, 
           y = density, color = color  ), 
       data = wq) +
  ylim(0.985,1.005) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1) 


```


Density and alcohol is more related in case of white wine. The second most 
significant variable for Red wine is volatile acidity. Looking at volatile 
acidity and alcohol relation.

``` {r, fig.width=6, fig.height=4, echo=FALSE, warning=FALSE, Bivariate_plot_5}

 ggplot(aes(x = alcohol, 
           y = volatile.acidity , color = color  ), 
       data = wq) +
  ylim(0.0,1.2) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1) 


```


The regression lines show that correlation of volatile acidity and alcohol is 
more in Red wine. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The correlation matrix shows correlation coefficients between each of the 
variables for both Red and White wines. It is interesting to note that the 
quality of wine is positively correlated with its alcohol level for both Red and
White wine. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Couple more results stand out:
1) the quality of Red wine has next significant 
correlation with volatile acidity while white wine has with density, chlorides,
and total sulphur dioxide.
2) residual sugar of white wine has strong negative correlation with the density
and alcohol, while of red wine does not have significant correlation.

### What was the strongest relationship you found?
Red Wine:
1) between alcohol and quality
2) between volatile acidity and quality
3) between fixed acidity and pH, density and citric acid.

White Wine:
1) between alcohol and quality
2) between density and quality
3) between density and alcohol and residual sugar.


# Multivariate Plots Section


Looking at Red and White wine indivudually next. Refering to the correlation 
matrix and comparing the two most significant variables determining the wine 
quality below.

``` {r  echo=FALSE, warning=FALSE, Multivariate_plots_1}

p1 <- ggplot( data= subset(wq, wq$color == "Red"),  aes(x=alcohol,
                            y = volatile.acidity, color = qualityF)) +
  ylim(0.0,1.2)+
  geom_point() +  
    geom_smooth(method = lm, se=TRUE) +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Alcohol [%]", y = "Volatile Acidity", color = "Quality")



p2 <- ggplot( data= subset(wq, wq$color == "White"),  aes(x=alcohol, 
                                            y = density, color = qualityF)) + 
   ylim(0.985,1.005) +
  geom_point() +  
    geom_smooth(method = lm, se=TRUE) +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Alcohol[%]", y = "Density", color = "Quality")

multiplot(p1, p2, cols = 1)

```

The scatterplot makes it easier to see the correlation of significant variables
and the quality of the wine separately for Red and White. Red wine has different
slopes for its regression lines for each wine quality while White wine has 
similar slopes for the three regression lines.

It will be interesting to see if we can predict the quality of fine depending on
these input variables. Plotting a simple desicion graph for both Red and White 
wine quaity. Used the two most significant correlated variables as inputs to the
decision tree.

``` {r fig.width=16, fig.height=8, echo = FALSE, Warning = FALSE, Multivariate_plot_2}


# preparing datasets for prediction algorithms
wqr_pred <- subset(wq, color == "Red",select= c(qualityF_sn, qualityF,
               fixed.acidity, volatile.acidity, citric.acid, residual.sugar, 
               chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, 
               pH, sulphates, alcohol))


wqw_pred <- subset(wq, color == "White",select= c(qualityF_sn, qualityF,
               fixed.acidity, volatile.acidity, citric.acid, residual.sugar, 
               chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, 
               pH, sulphates, alcohol))




# Create the tree for red wine.
  output.tree <- ctree(qualityF_sn ~ alcohol + volatile.acidity, 
  data = wqr_pred)
  
  plot(output.tree, main = "Decision tree: Red")
  
# Create the tree for white wine.  
  output.tree <- ctree(qualityF_sn ~ alcohol + density,  # + chlorides + total.sulfur.dioxide, 
  data = wqw_pred)

  plot(output.tree, , main = "Decision tree: White")
  
  

```


The output nodes have labels of e, f, p for excellent, fine and poor quality 
wines. Now using ctree and ramdom forest algorithm to predcit wine quality.

```{r echo=FALSE, warning=FALSE,  Multivariate_Plots_3}
set.seed(1)
sampr <- sample(nrow(wqr_pred), 0.6 * nrow(wqr_pred))
trainr <- wqr_pred[sampr, ]
testr <- wqr_pred[-sampr, ]

sampw <- sample(nrow(wqw_pred), 0.6 * nrow(wqw_pred))
trainw <- wqw_pred[sampw, ]
testw <- wqw_pred[-sampw, ]


# ctree prediction:
ct_modelr <- ctree(qualityF ~ alcohol + volatile.acidity , data=trainr)
ct_modelw <- ctree(qualityF ~ alcohol + density + chlorides + 
                     total.sulfur.dioxide, data=trainw)

predr <- predict(ct_modelr, newdata = testr)
tabler <- table(predr, testr$qualityF)
accuracyr_ct = sum(diag(tabler))/nrow(testr)

pred_w <- predict(ct_modelw, newdata = testw)
tablew <- table(pred_w, testw$qualityF)
accuracyw_ct = sum(diag(tablew))/nrow(testw)



#random forest prediction:
rf_modelr <- randomForest(qualityF ~ alcohol + volatile.acidity , data=trainr, 
                          importance=TRUE, ntree=500, mtry = 2)
rf_modelw <- randomForest(qualityF ~ alcohol + density + chlorides + 
                            total.sulfur.dioxide, data=trainw, 
                          importance=TRUE, ntree=500, mtry = 2)


predr <- predict(rf_modelr, newdata = testr)
tabler <- table(predr, testr$qualityF)
accuracyr_rf = sum(diag(tabler))/nrow(testr)

pred_w <- predict(rf_modelw, newdata = testw)
tablew <- table(pred_w, testw$qualityF)
accuracyw_rf = sum(diag(tablew))/nrow(testw)

accuracy <- matrix(c(accuracyr_ct,accuracyr_rf,accuracyw_ct,accuracyw_rf),
                   ncol=2)
colnames(accuracy) <- c('Red Wine', 'White Wine')
rownames(accuracy) <- c('Ctree', 'RandomForest')
accuracy.table <- as.table(accuracy)
accuracy.table

```

The table above shows the accuracy of both the algorithms in predicting both Red
and White wine quality. Random forest algorithm is better at predicting than 
ctree.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I analyzed decision trees in predicting wine quality in this section. Used the 
correlation coefficient matrix to determine the input to the algorithms.


### Were there any interesting or surprising interactions between features?

I used the correlation matrix to determine the inputs to prediction algorithm. 
The better performing algorithm has accuracy of 75% and 78% for Red and White 
wine respectively.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Created models from two different algorithms for predicting wine quality for 
each of the wine types. Computed the accuracy of the algorithm for comparison.

------

# Final Plots and Summary

### Plot One: Alcohol (%) in each quality level for Red and White Wines
```{r echo=FALSE, warning=FALSE, Plot_One}

p1 <- ggplot(aes(qualityF, 
            alcohol), 
        data = subset(wq, color == "Red",select= c(qualityF, alcohol)) )+
  geom_jitter( alpha = .3, color = 'blue')  +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Quality - Factored") + ylab("Alcohol [%]") + 
  scale_x_discrete(limits=c('poor','fine', 'excellent')) +
  geom_boxplot( alpha = .5,color = 'red')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

p2 <- ggplot(aes(qualityF, 
            alcohol), 
        data = subset(wq, color == "White",select= c(qualityF, alcohol)) ) +
  geom_jitter( alpha = .3, color = 'blue')  +
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
      xlab("Quality- Factored") + ylab("Alcohol [%]") + 
  scale_x_discrete(limits=c('poor','fine', 'excellent')) +
  geom_boxplot( alpha = .5,color = 'black')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 4)

multiplot(p1,p2, cols = 2)


```

### Description One
Figure 1 shows the the distribution of alchohol percentage with respect to wine
quality in both red and white wine. Wine quality improves with alcohol 
percentage. But, poor wines do have outliers with higher levels of alcohol. Mean
of each quality factor is comparable between wine types. For both Red and White,
fine wine's mean and median is same, while for poor and excellent wine's mean is
shifted.

### Plot Two: Significant variables correlation for each quality level for Red \
and White wine


```{r echo=FALSE, warning=FALSE, Plot_Two}


p1 <- ggplot( data= subset(wq, wq$color == "Red"),  aes(x=alcohol,
                            y = volatile.acidity, color = qualityF)) + 
  geom_point() +  
    geom_smooth(method = lm, se=TRUE) +
  ggtitle("Red") + theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Alcohol [%]", y = "Volatile Acidity [g / dm^3]", color = "Quality")



p2 <- ggplot( data= subset(wq, wq$color == "White"),  aes(x=alcohol, 
                                          y = density, color = qualityF)) + 
  geom_point() +   geom_smooth(method = lm, se=TRUE) +
  ylim(0.985, 1.005) +
    
  ggtitle("White") + theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Alcohol [%]", y = "Density [g / cm^3]", color = "Quality") 



multiplot(p1, p2, cols = 1)

```

### Description Two

Figure 2 shows relation between the two most significant variables correlated to 
wine quality for both Red and White wine. For Red wine plot shows vcolatile 
acidity vs alcohol and for White wine it shows density vs alcohol, for all types
of wine quality. Note, that in case of white wine for all wine quality types the
regression line has same slope. But, for Red wine the slope of the wine changes 
from positive to negative as wine quality improves. Which shows that there are 
some excellent wines with less alcohol percentage but more volatile acidity. But 
to make it conclusive we need more data points. 

### Plot Three: Simple decision tree for quality for Red and White wine

```{r fig.width=16, fig.height=8, echo=FALSE, warning=FALSE, Plot_Three}

# Create the tree for red wine.
  output.tree <- ctree(qualityF_sn ~ alcohol + volatile.acidity, 
  data = wqr_pred)
  

#png(file="animals45.png",width=400,height=350,res=45)
 plot(output.tree, main = "Decision tree: Red")
#dev.off()


  
# Create the tree for white wine.  
  output.tree <- ctree(qualityF_sn ~ alcohol + density,  
  data = wqw_pred)

  plot(output.tree, main = "Decision tree: White")
  
 #multiplot(p1, p2, cols = 1)

```

### Description Three
A simple decision matrix made with ctree algorithm is plotted in Figure 3 
separately for Red and White wine. Again the input variables used here where 
chosen based on the significance in the correlation coeeficient matrix. 

------

# Reflection

White and Red wine dataset was analyzed in the report. Initially explored one
varaible at a time and then moved to multiple variables. Since the variables had
numeric values I convereted couple of them into factors for exploration. The 
correlation matrix helped in determining the siginificant input variables. This
really helped further the analysis and give it a direction. I struggled a bit in 
couple of areas. One in deciding what kind of plots would display the varible
information efficiently and two making the plot pretty, with appropriate labels,
titles and figure size. A little bit of research, and playing around with ggplot
library functions helped finalizing the plots.

The main section in this report is the prediction of wine quality,in which the
significant varaibles were used as inputs. I used two algorithms for prediction 
and compared their accuracy. The highest accuracy computed was approx 80%. 
It would be interesting to look into other algorithms and perhaps explore if the
accuracy of these could be improved. Another area that I have missed to explore 
here is the relationship between other input variables, such as between fixed 
acidity and pH, density and citric acid in case of Red wine and between density 
and alcohol and residual sugar in case of White wine.


# References